Home

### 2025-07-19 about.qmd
Update website structure and content
Replaced the 'Research' section with a new 'About' page, detailing Professor Bin Yu's background and the Yu Group's research focus. Updated navigation links in _quarto.yml to reflect these changes, removing the old 'Research' link and adding 'About' and 'People' links. Enhanced the index.qmd layout for better presentation of recent publications and news, and added a responsive two-column layout in styles.css for improved readability.

---
title: Biography
---

# Biography

@@

@@

>>>>

---
title: About
---

# About Bin

@@

Professor Bin Yu was formally trained as a statistician, but her research interests and achievements extend beyond the realm of statistics. Together with her group, Bin Yu has leveraged new computational developments to solve important scientific problems by combining novel and often interpretable statistical machine learning approaches with the domain expertise of her many collaborators in neuroscience, genomics and precision medicine. She also develops relevant theory to understand random forests and deep learning for insight into and guidance for practice. Her work has been recognized by many awards. In particular, she was inducted to the National Academy of Sciences in 2014 and to the American Academy of Arts and Sciences in 2013. She delivered the IMS Wald Lectures and COPSS Distinguished Achievement and Award Lecture (DAAL, formerly Fisher Award and Lecture) in 2023.

@@

# About the Research Group

The [Yu Group](https://www.stat.berkeley.edu/~yugroup/) at Berkeley consists of 12-15 students and postdocs from Statistics and EECS. In the research group, we cultivate a strongly interdisciplinary and collaborative culture, solving data problems across fields such as neuroscience, genomics, remote sensing, and precision medicine.

Through these projects we have successfully [mapped a cell's destiny](https://newscenter.lbl.gov/2016/05/04/mapping-cells-destiny/) using spatial gene expression images of Drosophila embryos, we have characterized V4 neurons through [DeepTune](https://www.biorxiv.org/content/10.1101/465534v1) images, and we are currently seeking genomic markers of heart disease using UK Biobank data. Recently, my group and I have developed approaches for [predicting county-level COVID-19 death counts](https://covidseverity.com/) in an effort to support the non-profit, [Response4Life](https://response4life.org/), who are working towards distributing PPE across the country to those who need it most.

The Yu Group and I have also developed an array of statistical and machine learning methods inspired by our interdisciplinary projects, including [stability-driven nonnegative matrix factorization (staNMF)](https://github.com/Yu-Group/staNMF) for unsupervised learning, [iterative Random Forests (iRF)](https://github.com/Yu-Group/iterative-Random-Forest) and [signed iRF (s-iRF)](https://github.com/karlkumbier/iRF2.0) for discovering predictive and stable high-order (Boolean) interactions in supervised learning, [contextual decomposition (CD)](https://github.com/jamie-murdoch/ContextualDecomposition) and [aggregated contextual decomposition (ACD)](https://github.com/csinva/hierarchical-dnn-interpretations) for phrase or patch importance extraction from Deep Neural Networks (DNNs).

Research is supported in part by grants from NSF, NIH, the Weill Neurohub, and the Simons Foundation.

- [Yu Group at Berkeley](https://www.stat.berkeley.edu/~yugroup/) for current group members and current projects
- [Papers](https://binyu.stat.berkeley.edu/papers)
- [Code](https://www.stat.berkeley.edu/~yugroup/code.html)
- [Statistical Machine Learning at Berkeley](http://www.stat.berkeley.edu/~statlearning/)

# Vision

In 2014, I was elected to the National Academy of Sciences based on my statistical and scientific contributions, as well as my broad vision of data science best described in my article [Veridical Data Science](https://www.pnas.org/content/117/8/3920), written together with my former student Karl Kumbier. In this work, I introduced a framework based on three principles: predictability, computability and stability (abbreviated to PCS). This framework helps guide practitioners who solve domain data problems with data science tools to be creative in their analysis and properly validate their findings. I have written a book on the Veridical Data Science framework together with my former student [Rebecca Barter](https://www.rebeccabarter.com/). The book is being published by the MIT Press in 2024 with a free on-line copy available soon.

She and her team have developed the PCS framework for veridical data science (or responsible, reliable, and transparent data analysis and decision-making). PCS stands for predictability, computability and stability, and it unifies, streamlines, and expands on ideas and best practices of machine learning and statistics.

In order to augment empirical evidence for decision-making, they are investigating statistical machine learning methods/algorithms (and associated statistical inference problems) such as dictionary learning, non-negative matrix factorization (NMF), EM and deep learning (CNNs and LSTMs), and heterogeneous effect estimation in randomized experiments (X-learner). Their recent algorithms include staNMF for unsupervised learning, iterative Random Forests (iRF) and signed iRF (s-iRF) for discovering predictive and stable high-order interactions in supervised learning, next generation tree-based methods (e.g. fast and interpretable greedy-tree sums (FIGS) and hierarchical shrinked (HS) trees, and RF+ ), contextual decomposition (CD), aggregated contextual decomposition (ACD), and adaptive wavelet distillation (AWD) for interpretation of Deep Neural Networks (DNNs).

### 2025-07-19 index.qmd
Update index.qmd
Update website structure and content
Replaced the 'Research' section with a new 'About' page, detailing Professor Bin Yu's background and the Yu Group's research focus. Updated navigation links in _quarto.yml to reflect these changes, removing the old 'Research' link and adding 'About' and 'People' links. Enhanced the index.qmd layout for better presentation of recent publications and news, and added a responsive two-column layout in styles.css for improved readability.

## Welcome

Professor Bin Yu is the head of the [Yu Group](https://www.stat.berkeley.edu/~yugroup/) at Berkeley, which consists of 12-15 students and postdocs from Statistics and EECS. She was formally trained as a statistician, but her research interests and achievements extend beyond the realm of statistics. Together with her group, Bin Yu has leveraged new computational developments to solve important scientific problems by combining novel and often interpretable statistical machine learning approaches with the domain expertise of her many collaborators in neuroscience, genomics and precision medicine. She also develops relevant theory to understand random forests and deep learning for insight into and guidance for practice. Her work has been recognized by many awards. In particular, she was inducted to the National Academy of Sciences in 2014 and to the American Academy of Arts and Sciences in 2013. She delivered the IMS Wald Lectures and COPSS Distinguished Achievement and Award Lecture (DAAL, formerly Fisher Award and Lecture) in 2023.

She and her team have developed the PCS framework for veridical data science (or responsible, reliable, and transparent data analysis and decision-making). PCS stands for predictability, computability and stability, and it unifies, streamlines, and expands on ideas and best practices of machine learning and statistics.

In order to augment empirical evidence for decision-making, they are investigating statistical machine learning methods/algorithms (and associated statistical inference problems) such as dictionary learning, non-negative matrix factorization (NMF), EM and deep learning (CNNs and LSTMs), and heterogeneous effect estimation in randomized experiments (X-learner). Their recent algorithms include staNMF for unsupervised learning, iterative Random Forests (iRF) and signed iRF (s-iRF) for discovering predictive and stable high-order interactions in supervised learning, next generation tree-based methods (e.g. fast and interpretable greedy-tree sums (FIGS) and hierarchical shrinked (HS) trees, and RF+ ), contextual decomposition (CD), aggregated contextual decomposition (ACD), and adaptive wavelet distillation (AWD) for interpretation of Deep Neural Networks (DNNs).

## My vision for data science - papers & talks

- [Rome Workshop on Veridical Data Science, June 20, 2025](https://www.integreat.no/events/public-events/workshops/veridical-data-science.html) (New)
- [Veridical data science and medical foundation models](https://arxiv.org/abs/2409.10580) on arXiv by Alaa and Yu.
- [VDS book review](https://hdsr.mitpress.mit.edu/pub/5322mn1e/release/3) by Yuval and Yoav Benjamini in Harvard Data Science Review (HDSR)-- "pedagogical excellence, diverse examples, and projects make Veridical Data Science a suitable textbook for students of all levels, in addition to being a valuable resource for data scientists in general" (review summary in HDSR Editor-in-Chief's note), June, 2024.
- [Inaugural Berkeley-Stanford Workshop on Veridical Data Science at UC Berkeley](https://na.eventscloud.com/website/69057/) (May 31, 2024) ([talk videos available](https://www.youtube.com/playlist?list=PLKW2Azk23ZtSrg4jwLSioaV1bwmFnQ1Bx))
- My co-author [Rebecca Barter](https://rebeccabarter.com/) and I are thrilled to announce the [online release](http://vdsbook.com/) of our MIT Press book ["Veridical Data Science: The Practice of Responsible Data Analysis and Decision Making"](https://mitpress.mit.edu/9780262049191/veridical-data-science/), an essential source for producing trustworthy data-driven results (Feb. 28, 2024).
- ["Veridical data science towards trustworthy AI"](https://ww2.amstat.org/meetings/jsm/2023/webcasts/index.cfm) (Talk video of COPSS DAAL (formerly Fisher Award and Lecture), scroll down), Aug. JSM in Toronto, 2023.
- [What is uncertainty in today's practice of data science?](https://www.sciencedirect.com/science/article/pii/S030440762300235X?dgcid=author) (Yu, 2023). Journal of Econometrics.
- [Building trust in medical AI algorithms with veridical data science](https://link.springer.com/epdf/10.1007/s13218-023-00803-y?sharing_token=ebI2N15qM6XZUegMth-Quve4RwlQNchNByi7wbcMAY7HhSJ5V2qZIFNLqxQFZ-zTvebYYHuYBEDcXzQl7eRE1JVsvjcy_eFebw0xZW9UU9CqWdFLLGd22GvHmjiyaKtyaRMQToSpZQruTneGX1KD5rV1I5kdcOMfqUq47TrVW5E%3D) (interview of Bin Yu by Dr. Merle Behr for the German scientific journal [KI - Künstliche Intelligenz](https://www.springer.com/journal/13218)), 2023
- [Veridical data science](https://www.pnas.org/content/117/8/3920) (PCS framework: [v-flow code](https://github.com/Yu-Group/veridical-flow) and [documentation template](https://yu-group.github.io/vdocs/PCSDoc-Template.html)), PNAS, 2020 ([QnAs with Bin Yu](https://www.pnas.org/content/117/8/3893))
- [Breiman Lecture (video) at NeurIPS "Veridical data Science"](https://slideslive.com/38922599/veridical-data-science) ([PCS framework](https://www.pnas.org/content/117/8/3920) and iterative random forests ([iRF](https://www.pnas.org/content/115/8/1943))), 2019; [updated slides](https://www.stat.berkeley.edu/~binyu/ps/papers2020/Breiman19-NeurIPS-yu.pdf), 2020
- [Stability](https://projecteuclid.org/euclid.bj/1377612862), Bernoulli, 2013
- [Stability expanded, in reality](https://hdsr.mitpress.mit.edu/pub/ekrhsui8/release/2), Harvard Data Science Review (HDSR), 2020.
- [Data science process: one culture.](https://www.tandfonline.com/doi/abs/10.1080/01621459.2020.1762615?journalCode=uasa20) JASA, 2020.
- [Minimum information about clinical artificial intelligence modeling: the MI-CLAIM checklist](https://www.nature.com/articles/s41591-020-1041-y), Nature Medicine, 2020.
- [Definitions, methods and applications in interpretable machine learning](https://www.pnas.org/content/116/44/22071), PNAS, 2019
- [Data wisdom for data science (blog)](http://www.odbms.org/2015/04/data-wisdom-for-data-science/), 2015
- [IMS Presidential Address "Let us own data science"](https://imstat.org/2014/10/01/ims-presidential-address-let-us-own-data-science/), IMS Bulletin, 2014
- [Embracing statistical challenges in the IT age](https://www.tandfonline.com/doi/abs/10.1198/004017007000000254), Technometrics, 2007

## In the news

- IMS Wald [Lecture I](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1029) and [Lecture II](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1031), and [COPSS Distinguished Award and Lecture (DAAL)](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1374) delivered at [Joint Statistical Meeting (JSM)](https://ww2.amstat.org/meetings/jsm/2023/) in Toronto, August 2023
- UC Berkeley team part of the new NSF AI Center for Cybersecurity [ACTION](https://action.ucsb.edu/) led by UCSB.
- [CDSS news: Statistics-Computer Science team reflects on tackling covid outbreaks](https://data.berkeley.edu/news/statistics-computer-sciences-team-reflects-tackling-covid-outbreaks), May, 2022
- Honorary Doctorate, University of Lausanne (UNIL) (Faculty of Business and Economics), June 4, 2021 ([Interview of Bin Yu](https://www.youtube.com/watch?v=1JRGPGhcYzU) by journalist Nathalie Randin, with an introduction by Dean Jean-Philippe Bonardi of UNIL in French ([English translation](https://www.stat.berkeley.edu/~binyu/Intro-DeanBonardi-UNIL.pdf)))
- [CDSS news on our PCS framework: "A better framework for more robust, trustworthy data science"](https://data.berkeley.edu/people/bin-yu), Oct. 2020
- [UC Berkeley to lead $10M NSF/Simons Foundation program to investigate theoretical underpinnings of deep learning](https://data.berkeley.edu/news/uc-berkeley-lead-10m-nsfsimons-foundation-program-investigate-theoretical-underpinnings-deep), Aug. 25, 2020
- [Curating COVID-19 data repository and forecasting county-level death counts in the US](https://engineering.berkeley.edu/news/2020/04/getting-the-right-equipment-to-the-right-people/), 2020
- [Interviewed by PBS Nova about AlphaZero](https://www.pbs.org/wgbh/nova/article/google-alphazero-artificial-intelligence/), 2018
- [Mapping a cell's destiny](https://newscenter.lbl.gov/2016/05/04/mapping-cells-destiny/), 2016
- [Seeking Data Wisdom](https://vcresearch.berkeley.edu/news/seeking-data-wisdom), 2015
- [Member, National Academy of Sciences](http://www.dcd.zju.edu.cn/Hsu-conference/pr.htm), 2014
- [Fellow, American Academy of Arts and Sciences](http://www.dcd.zju.edu.cn/Hsu-conference/pr.htm), 2013
- [One of the 50 best inventions of 2011 by Time Magazine](http://content.time.com/time/magazine/article/0,9171,2099708,00.html), 2011
- [The Economist Article](http://www.economist.com/node/21534748), 2011
- [ScienceMatters @ Berkeley. Dealing with Cloudy Data](https://www.stat.berkeley.edu/~binyu/ps/dealingwithcloudydata.pdf), 2004

>>>>

<div class="two-column-container">

<div class="column">
## Recent Publications & Talks

- [New paper: PCS workflow for veridical data science in the age of AI (Z. Rewolinski and B. Yu, 2025)](https://www.stat.berkeley.edu/~yugroup/pubs/pcs_workflow-6.pdf)
- [Veridical data science and medical foundation models](https://arxiv.org/abs/2409.10580) on arXiv by Alaa and Yu.
- [VDS book review](https://hdsr.mitpress.mit.edu/pub/5322mn1e/release/3) by Yuval and Yoav Benjamini in Harvard Data Science Review (HDSR)-- "pedagogical excellence, diverse examples, and projects make Veridical Data Science a suitable textbook for students of all levels, in addition to being a valuable resource for data scientists in general" (review summary in HDSR Editor-in-Chief's note), June, 2024.
- ["Veridical data science towards trustworthy AI"](https://ww2.amstat.org/meetings/jsm/2023/webcasts/index.cfm) (Talk video of COPSS DAAL (formerly Fisher Award and Lecture), scroll down), Aug. JSM in Toronto, 2023.
- [Veridical data science](https://www.pnas.org/content/117/8/3920) (PCS framework: [v-flow code](https://github.com/Yu-Group/veridical-flow) and [documentation template](https://yu-group.github.io/vdocs/PCSDoc-Template.html)), PNAS, 2020 ([QnAs with Bin Yu](https://www.pnas.org/content/117/8/3893))
- My co-author [Rebecca Barter](https://rebeccabarter.com/) and I are thrilled to announce the [online release](http://vdsbook.com/) of our MIT Press book ["Veridical Data Science: The Practice of Responsible Data Analysis and Decision Making"](https://mitpress.mit.edu/9780262049191/veridical-data-science/), an essential source for producing trustworthy data-driven results (Feb. 28, 2024).

[View all papers →](papers.qmd)

</div>

<div class="column">
## Recent News

- [Veridical Data Science in Biology on July 11, 2025 at UC Berkeley (Submission deadline July 4)](https://www.eventbrite.com/e/veridical-data-science-for-biology-2025-tickets-1384456339179)
- [Rome Workshop on Veridical Data Science, June 20, 2025](https://www.integreat.no/events/public-events/workshops/veridical-data-science.html)
- IMS Wald [Lecture I](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1029) and [Lecture II](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1031), and [COPSS Distinguished Award and Lecture (DAAL)](https://ww2.aievolution.com/JSMAnnual/index.cfm?do=ev.viewEv&ev=1374) delivered at [Joint Statistical Meeting (JSM)](https://ww2.amstat.org/meetings/jsm/2023/) in Toronto, August 2023
- [Inaugural Berkeley-Stanford Workshop on Veridical Data Science at UC Berkeley](https://na.eventscloud.com/website/69057/) (May 31, 2024) ([talk videos available](https://www.youtube.com/playlist?list=PLKW2Azk23ZtSrg4jwLSioaV1bwmFnQ1Bx))
- UC Berkeley team part of the new NSF AI Center for Cybersecurity [ACTION](https://action.ucsb.edu/) led by UCSB.
- [CDSS news: Statistics-Computer Science team reflects on tackling covid outbreaks](https://data.berkeley.edu/news/statistics-computer-sciences-team-reflects-tackling-covid-outbreaks), May, 2022
- Honorary Doctorate, University of Lausanne (UNIL) (Faculty of Business and Economics), June 4, 2021 ([Interview of Bin Yu](https://www.youtube.com/watch?v=1JRGPGhcYzU) by journalist Nathalie Randin, with an introduction by Dean Jean-Philippe Bonardi of UNIL in French ([English translation](https://www.stat.berkeley.edu/~binyu/Intro-DeanBonardi-UNIL.pdf)))
- [CDSS news on our PCS framework: "A better framework for more robust, trustworthy data science"](https://data.berkeley.edu/people/bin-yu), Oct. 2020
- [UC Berkeley to lead $10M NSF/Simons Foundation program to investigate theoretical underpinnings of deep learning](https://data.berkeley.edu/news/uc-berkeley-lead-10m-nsfsimons-foundation-program-investigate-theoretical-underpinnings-deep), Aug. 25, 2020

</div>

</div>
